Future cosmic microwave background delensing with galaxy surveys.

### 2017-09-28 Notes/galaxy_delensing.tex

\title{Future cosmic microwave background delensing with galaxy surveys.}       
\author{A. Manzotti} 
\email{manzotti.alessandro@gmail.com}
\affiliation{Kavli Institute for Cosmological Physics, University of Chicago}
\affiliation{Department of Astronomy and Astrophysics, University of Chicago, 5640 South Ellis Avenue, Chicago, IL 60637, USA}
\date{\today}
\begin{abstract}

@@

\begin{figure}[htbp]
\begin{center}
\includegraphics[scale=1.]{../images/BB_res_ell2.pdf}
\caption{Here we illustrate the effect of delensing on the B-mode power spectrum. The red solid line corresponds to the fiducial lensing B-mode component of the signal while the dashed blue line corresponds to the inflationary one for the higher amplitude allowed by current experiments.
These are compared with the residual power left after delensing using some of the LSS tracers described in \refsec{model}. 
The rapid improvement in the level of instrumental noise (dashed curves for SPTPol and CMB S4) will require a high delensing efficiency to exploit these experiments fully.}
\label{fig:bbplot}
\end{center}
\end{figure}

@@

% ===================
\begin{figure}[htbp]
\begin{center}
\includegraphics[scale=1.]{../images/compare_kernel.pdf}
\caption{Kernel Comparison: Comparison of the different kernels as a function of redshift for some of the tracers used in this analysis. The bigger the overlap with the CMB lensing kernel the better the reconstruction of the lensing potential will be leading to a higher delensing efficiency. }
\label{fig:kernel}
\end{center}
\end{figure}
% ===================

@@

% ===================
\begin{figure}[htbp]
\begin{center}
\includegraphics[scale=1.]{../images/compare_kernel_tomo.pdf}
\caption{Kernel Comparison with tomographic bins: Comparison of the 10 LSST  tomographic bins together with the full LSST survey and the CMB lensing kernel. 
Compared to a full survey approach, tomographic binning allows to optimally weight different bins according to their cross-correlation with the CMB lensing. This leads to a better delensing efficiency. }
\label{fig:kerneltomo}
\end{center}
\end{figure}
% ===================

@@

\begin{figure}[htbp]
\begin{center}
\includegraphics[scale=1.]{../images/B_res_bin.pdf}
\caption{Tomographic bins improve the cross-correlation of galaxy surveys with CMB lensing. Here we show the cross-correlation coefficient (\refeq{rho1}) as a function of angular scale for full surveys (solid lines) and tomographically binned surveys (dot-dashed line).}
\label{fig:rho-tomo}
\end{center}
\end{figure}

@@

In this section, we forecast the expected delensing efficiency and the relative importance of galaxy tracers for delensing in current and future experiments. We will use the Fisher information matrix to quantify the delensing efficiency as the improvement in the constraint of two inflationary parameters: the tensor to scalar ratio $r$ and the tensor tilt $n_{T}$.
We assume a CMB experimental scenario composed of a high-resolution CMB experiment which defines the ability to internally reconstruct the potential $\kappa$, together with a low-noise, low-resolution experiment that measured the B-modes that will be delensed and used to constrain the inflationary parameters.
We focus on three distinct scenarios: the current stage, a third generation stage (3G)  and finally the futuristic CMB Stage 4.

@@

It can be seen in \refeq{fisher} that removing the lensing contribution will improve parameter constraints.
The parameter uncertainties are inversely proportional to the covariance of the measurement. Since the lensing B-modes power $C_{{l}}^{BB,\mathrm{lens}}$ is a substantial component of the covariance, removing part of it through delensing will reduce the parameter statistical error.
We define the error after delensing as $\sigma^{\rm{del}}(p)$.
This is computed from \refeq{fisher} substituting $C_{{l}}^{BB,\mathrm{lens}}$ with the residual power after delensing $C_{{l}}^{BB,\mathrm{res}}$ in the covariance matrix defined in \refeq{cova}.
We define the improvement as the ratio of the constraints before and after delensing: $\alpha_{r} = \sigma^{\rm{del}}(r)/\sigma(r)$ and $\alpha_{n_{t}} = \sigma^{\rm{del}}(n_{t})/\sigma(n_{t})$. The fiducial values used are described in \refsec{del_res}. Given a fiducial value for $r$, the value of $n_{t}$ is fixed imposing the consistency relation $n_t=-r/8$ \cite{LIDDLE1992391}.

@@

Recently delensing has been performed for the first time on data using both CIB maps \cite{larsen:2016a,manzotti:2017} and internal CMB lensing maps \cite{carron:2017} as large scale structure tracers.
Here we discuss the improvement that can be obtained by combining these and other currently available tracers (see \cite{yu:2017} for publicly accessible multi-tracers data-products).

On the CMB side, we will combine a BICEP-Keck like \textit{deep} CMB experiment with an overlapping \textit{higher resolution} experiment. For the deep experiment we assume an instrumental noise equal to \textcolor{red}{$3\mu$K-arcmin}, a beam of \textcolor{red}{30 arcmin} and an angular scale range of $50<\ell<500$.
We assume the CMB lensing reconstruction is  performed by the higher resolution CMB experiment. We explore two possibilities.
First, we test the efficiency with an internal reconstruction of CMB lensing performed by the Planck collaboration \cite{planck-collaboration:2016a}. We compute the noise in the CMB lensing map in \refeq{noisekappa} using the noise curves publicly available
\footnote{\url{https://wiki.cosmos.esa.int/planckpla2015/index.php/Specially\_processed\_maps}}.
Then, we test a second scenario with a ground-based experiment with noise levels consistent with SPT-Pol: \textcolor{red}{$9.4\mu$K-arcmin} in polarization with a beam equal to \textcolor{red}{1.2 arcmin}. In this case we set the largest achievable scale at $\ell_{min}=300$. However, the results are robust against this choice since the neglected scales contribute only moderately to the reconstruction of the CMB lensing potential.

@@

\begin{figure}[htbp]
\begin{center}
\includegraphics[scale=1.]{../images/actual_scenario_planck.pdf}
\caption{Correlation factor $\rho$ with the lensing potential $\kappa$ as a function of the angular scale $\ell$ for current generation tracers.
Here we show both current galaxy survey and internally reconstructed $\phi$ CMB lensing potential. 
The dashed curve highlights which scales contribute to the $\ell<100$ B-mode power and are thus most useful to delens. It corresponds to (with arbitrary scale) $C^{\kappa \kappa} \times \frac{\partial C^{BB}}{\partial{C^{\kappa \kappa}}}$}
\label{fig:corrnow-planck}
\end{center}
\end{figure}

\begin{figure}[htbp]
\begin{center}
\includegraphics[scale=1.]{../images/actual_scenario.pdf}
\caption{Correlation factor $\rho$ with the lensing potential $\kappa$ as a function of the angular scale $\ell$ for current generation tracers. 
This figure is the same as \reffig{corrnow-planck} but using SPTPol instead of Planck as the high-res CMB experiment performing the internal reconstruction.
}
\label{fig:corrnow}
\end{center}
\end{figure}

Using these correlation levels we can  compute the residual B-mode power after delensing using \refeq{Bres}, and test the consequent improvement on parameter constraints with \refeq{fisher}. 
For current surveys we only consider $p=r$ since the broad beam of the deep BK-like survey does not provide enough leverage to constrain the shape of the B-mode spectra. We still let $n_{t}$ to vary in our fisher calculation even if it is not very degenerate with $r$.
We first compute the improvement with a fiducial value of $r_{\rm{fid}}=0$.
The results are summarized in \reftab{current} and \reftab{current-spt} for the Planck and SPTPol case respectively.

The lensing reconstruction noise in Planck is such that internal delensing can only remove $8\%$ of the power, improving the constraints on $r$ by a factor of 1.06 even in the ideal scenario of no instrumental noise in the B-modes. A similar performance is achieved with WISE. DES can do better, removing a level of 14\% of power. However, it covers a smaller fraction of the sky than the two previous probes. As shown in previous works \cite{sherwin:2015}, the best tracer is the CIB with a 27\% reduction in power.
Nonetheless, it is still worthwhile considering all the tracers. Indeed a multi tracer-approach can bring the removed power on the overlapping area from 27\% of the CIB alone to 36\% using all the LSS (LSS-S2) and 41\% when Planck lensing reconstruction is added to the set.
However, if we consider a realistic level of noise in the measured B-modes, this level of residual power $C_{{l}}^{BB,\mathrm{res}}$ only leads to a 30\% improvement in $\sigma(r)$ for the null hypothesis case of $r=0$. Furthermore, as explained in \refsec{fisher}, this is only an upper limit and will be even smaller when, for example, non-perfect foreground cleaning is considered. Indeed current experiments are still significantly limited by foregrounds and instrumental noise together and not only by lensing.

@@

\begin{table}
\caption{$\alpha(r)$: Improvements on $\sigma(r)$ due to delensing for current generation and Planck lensing reconstruction.
LSS-S2 corresponds to the combination of all the available LSS tracers.
The values in parenthesis in the first column correspond to the fraction of lensing B-mode power removed using each LSS tracer. 
The values in the other columns correspond to ratio of the error before and after delensing.}
  \vspace{0.2cm}
  \begin{tabular}{|c | c | c | c|}
\hline
Surveys & $\alpha_{r=0}, N^{B}_{\ell}=0$ & $\alpha_{r=0}$ & $\alpha_{r=0.12}$ \\ \hline \hline
WISE (6\%)& 1.09& 1.05 & 1.01 \\ \hline
DES (14\%) &1.2 & 1.10  & 1.03 \\ \hline
CIB (27\%) &1.45& 1.21  & 1.07\\ \hline
LSS-S2 (36\%) &1.7& 1.29  & 1.09\\ \hline
CMB Planck (8\%) &1.06& 1.05  & 1.01 \\ \hline
LSS-S2 +CMB (41\%)& 1.77& 1.31& 1.10  \\ \hline
\end{tabular}
\label{tab:current}
\end{table}

@@

 \begin{table}
\caption{$\alpha(r)$: Current generation improvements on $\sigma(r)$ due to delensing. We use the same LSS tracers as \reftab{current} but with the internal lensing reconstruction performed by SPTPol.}
  \vspace{0.2cm}
  \begin{tabular}{|c | c | c | c|}
\hline
Surveys &$\alpha_{r=0}, N^{B}_{\ell}=0$& $\alpha_{r=0}$ & $\alpha_{r=0.12}$ \\ \hline \hline
SPTPol (35\%) &1.46& 1.27  & 1.08 \\ \hline
LSS-S2 +SPTPol (56\%) & 2.18 & 1.57 & 1.16  \\ \hline

@@

\subsection{CMB-S3 Era}\label{sec:S3}
CMB polarization measurements are rapidly improving. Indeed the next generation of ground-based telescopes has been already deployed, and data are currently being taken.
As we did for Stage-2, we model the CMB S3 as two different overlapping experiments that will be able to combine their measurement.
The two experiments combined will have a level of noise of \textcolor{red}{$3\mu$K-arcmin} and a \textcolor{red}{1 arcmin} beam. This level of noise is also assumed for the internal noise reconstruction.
With this experimental setup, we can push the angular scale range used to calculate the Fisher matrix to $50<\ell<3000$ even if most of the high $\ell$ scale do not contribute to the constraints.
Not only will CMB experiments improve shortly: eventually DESI will start taking data. For this reason, we add DESI to the LSS tracers used in \refsec{del_res}.

@@

\begin{figure}[htbp]
\begin{center}
\includegraphics[scale=1.]{../images/S3_scenario.pdf}
\caption{Correlation factor $\rho$ with the lensing potential $\kappa$ as a function of the angular scale $\ell$. This figure is the same as \reffig{corrnow} but with a stage-3 CMB experiment performing the internal reconstrcution and with DESI added to the stage-2 galaxy surveys (LSS-S2). }
\label{fig:corrS3}
\end{center}
\end{figure}

@@

\subsection{CMB-S4 Era}
Having as a major goal the detection of inflationary B-modes, CMB data will continue improving even after Stage-3.
An ambitious program for a Stage-4 ground CMB experiment is in the planning phase \cite{abazajian:2016}. 
Moreover, satellite and balloon CMB experiments have been proposed and have the potential to extend the accessible B-mode measurements to the largest scales. Given the unprecedented low level of noise of these experiment, delensing will be even more important than previous generations.
Following \cite{abazajian:2016}, here we assume a CMB-S4 ground experiment with a level of noise of \textcolor{red}{$1\mu$K-arcmin} and a \textcolor{red}{1 arcmin} beam.
Furthermore, by the mid-2020s, several next-generation LSS surveys will be online. 
Here, as an example of future optical galaxy surveys, we add LSST \cite{lsst-science-collaboration:2009} to the CMB lensing tracers. 
We do not consider other surveys but we test that other experiments like Euclid \cite{laureijs:2011} and WFIRST \cite{spergel:2013} have similar delensing performance. 
Even if these LSS surveys will observe an overlapping part of the sky and similar modes, because of their different strengths and weaknesses, it will still be important to combine them efficiently.
We also consider a radio continuum survey modeled following the SKA specifications. Radio continuum observations of the 21 cm line are in their early stage, and several experimental and data-analysis challenges need to be overcome. 
However, this technique has the potential to map the LSS with relatively low noise up to redshift $z=6$. Here we considered just two possible detection thresholds at 1 GHz (flux cut): 10 $\mu$Jy and 0.1 $\mu$Jy, which are representatives of the upper and lower limit levels for a SKA phase 1 (SKA1) and a following SKA phase 2 (SKA2) survey.
Given the importance of delensing for Stage-4 experiments, here we combine radio and optical survey to test their delensing efficiency.

The correlation factor with CMB lensing for Stage-4 experiments is shown in \reffig{corrs4_10}  and in \reffig{corrs4_01} for SKA1 and SKA2 respectively.
%=======
\begin{figure}[htbp]
\begin{center}
\includegraphics[scale=1.]{../images/S4_scenario_ska10.pdf}
\caption{Correlation factor $\rho$ with the lensing potential $\kappa$ as a function of the angular scale $\ell$. This figure is the same as \reffig{corrnow} but with a stage-4 CMB experiment performing the internal reconstrcution and with LSST and SKA added to the stage-2 and stage-3 galaxy surveys (LSS-S3). }
\label{fig:corrs4_10}
\end{center}
\end{figure}
%=======
\begin{figure}[htbp]
\begin{center}
\includegraphics[scale=1.]{../images/S4_scenario_ska01.pdf}
\caption{Correlation factor $\rho$ with the lensing potential $\kappa$ as a function of the angular scale $\ell$. This figure is the same as \reffig{corrnow} but with a stage-4 CMB experiment performing the internal reconstrcution and with LSST and SKA added to the stage-2 and stage-3 galaxy surveys (LSS-S3). }
\label{fig:corrs4_01}
\end{center}
\end{figure}
%=======
A tomographically binned LSST-like experiment will be a very efficient CMB lensing tracer. Indeed it will be more than 70\% correlated with CMB lensing for all the scales $\ell<600$, a performance similar to a Stage-3 CMB internal reconstruction.
A very similar level of efficiency will be achieved by a SKA1-like experiment.
With SKA and LSST, for the first time we will have LSS tracers with a higher delensing efficiency than the CIB.
Despite the improvement in galaxy surveys, the CMB internal reconstruction will still be the main source of delensing. Having a perfect kernel overlap with the true lensing potential it will benefit from the very low level of noise of CMB S4 experiments.
\reffig{corrs4_10} shows that a CMB-S4 experiment will be able to internally reconstruct the lensing potential at more than 90\% up to $\ell=400$.
At this level, CMB S4 internal delensing will not be limited by noise but by small secondary effects like foregrounds contamination, filtered modes etc.
Finally, \reffig{corrs4_01} shows that a very futuristic SKA-2 experiment will have a very high correlation with CMB lensing, being more than 80\% correlated at all the scales. This performance is comparable to a CMB-S4 internal reconstruction.

@@

%=================================================================
\section{Acknowledgments}
We thank S. Dodelson, S. Passaglia, K. Story and K. Wu for useful discussion and their review of the draft. 
Furthermore we thank D. Alonso for his feedback on radio continuum surveys.

>>>>

\title{Future cosmic microwave background delensing with galaxy surveys.}       
\author{A. Manzotti} 
\email{manzotti.alessandro@gmail.com}
\affiliation{Kavli Institute for Cosmological Physics, University of Chicago}
\affiliation{Department of Astronomy and Astrophysics, University of Chicago, 5640 South Ellis Avenue, Chicago, IL 60637, USA}
\date{\today}

\begin{abstract}

@@

\begin{figure}[htbp]
\begin{center}
\includegraphics[scale=1.]{../images/BB_res_ell2.pdf}
\caption{Here we illustrate the effect of delensing on the B-mode power spectrum. The orange solid line corresponds to the fiducial lensing B-mode component of the signal while the dashed blue line corresponds to the inflationary one for the higher amplitude allowed by current experiments.
These are compared with the residual power left after delensing using some of the LSS tracers described in \refsec{model}. 
The rapid improvement in the level of instrumental noise (dashed curves for SPTPol and CMB S4) will require a high delensing efficiency to exploit these experiments fully.}
\label{fig:bbplot}
\end{center}
\end{figure}

@@

% ===================
\begin{figure}[htbp]
\begin{center}
\includegraphics[scale=1.]{../images/compare_kernel.pdf}
\caption{Kernel Comparison: Comparison of the different kernels as a function of redshift for some of the tracers used in this analysis. The bigger the overlap with the CMB lensing kernel the better the reconstruction of the lensing potential is thus leading to a higher delensing efficiency. }
\label{fig:kernel}
\end{center}
\end{figure}
% ===================

@@

% ===================
\begin{figure}[htbp]
\begin{center}
\includegraphics[scale=1.]{../images/compare_kernel_tomo.pdf}
\caption{Comparison of kernels of the 10 LSST tomographic bins together with the full LSST survey and the CMB lensing kernel. 
Compared to a full survey approach, tomographic binning allows to optimally weight different bins according to their cross-correlation with the CMB lensing. This leads to a better delensing efficiency. }
\label{fig:kerneltomo}
\end{center}
\end{figure}
% ===================

@@

\begin{figure}[htbp]
\begin{center}
\includegraphics[scale=1.]{../images/B_res_bin.pdf}
\caption{Tomographic bins improve the cross-correlation of galaxy surveys with CMB lensing. Here we show the cross-correlation coefficient (\refeq{rho1}) as a function of angular scale for full surveys (solid lines) and tomographically binned surveys (dot-dashed line). For details about the binning, see \refsec{tomo-bin} and in particular \refeq{zbin}. }
\label{fig:rho-tomo}
\end{center}
\end{figure}

@@

In this section, we forecast the expected delensing efficiency and the relative importance of galaxy tracers for delensing in current and future experiments. We will use the Fisher information matrix to quantify the delensing efficiency as the improvement in the constraint of two inflationary parameters: the tensor to scalar ratio $r$ and the tensor tilt $n_{T}$.
We assume a CMB experimental scenario composed of a high-resolution CMB experiment which defines the ability to internally reconstruct the potential $\kappa$, together with a low-noise, low-resolution experiment that measured the B-modes that will be delensed and used to constrain the inflationary parameters.
Even if there is not a definite distinction between different stages, we focus on three distinct scenarios: the current stage, a third generation stage (3G)  and finally the futuristic CMB Stage 4.

@@

It can be seen in \refeq{fisher} that removing the lensing contribution will improve parameter constraints.
The parameter uncertainties are inversely proportional to the covariance of the measurement. Since the lensing B-modes power $C_{{l}}^{BB,\mathrm{lens}}$ is a substantial component of the covariance, removing part of it through delensing will reduce the parameter statistical error.
We define the error after delensing as $\sigma^{\rm{del}}(p)$.
This is computed from \refeq{fisher} substituting $C_{{l}}^{BB,\mathrm{lens}}$ with the residual power after delensing $C_{{l}}^{BB,\mathrm{res}}$ in the covariance matrix defined in \refeq{cova}.
We define the improvement as the ratio of the constraints before and after delensing: $\alpha_{r} = \sigma^{\rm{del}}(r)/\sigma(r)$ and $\alpha_{n_{t}} = \sigma^{\rm{del}}(n_{t})/\sigma(n_{t})$. The fiducial values used are described in \refsec{del_res}. Given a fiducial value for $r$, the value of $n_{t}$ is fixed imposing the consistency relation $n_t=-r/8$ \cite{LIDDLE1992391}.
In order to minimize the degeneracy between $r$ and $n_{t}$ \cite{abazajian:2016}, we choose as a pivot scale $k_{t}=0.01 $ Mpc$^{-1}$ .

@@

Recently delensing has been performed for the first time on data using both CIB maps \cite{larsen:2016a,manzotti:2017} and internal CMB lensing maps \cite{carron:2017} as large scale structure tracers.
Here we discuss the improvement that can be obtained by combining these and other currently Stage-2 available tracers (see \cite{yu:2017} for publicly accessible multi-tracers data-products).

On the CMB side, we will combine a BICEP-Keck like \textit{deep} CMB experiment with an overlapping \textit{higher resolution} experiment. For the deep experiment we assume an instrumental noise equal to $3\mu$K-arcmin, a beam of 30 arcmin and an angular scale range of $50<\ell<500$.
We assume the CMB lensing reconstruction is  performed by the higher resolution CMB experiment. We explore two possibilities.
First, we test the delensing efficiency with an internal reconstruction of CMB lensing performed by the Planck collaboration \cite{planck-collaboration:2016a}. We compute the noise in the CMB lensing map in \refeq{noisekappa} using the noise curves publicly available
\footnote{\url{https://wiki.cosmos.esa.int/planckpla2015/index.php/Specially\_processed\_maps}}.
Then, we test a second scenario with a ground-based experiment characterized by noise levels consistent with SPT-Pol \cite{henning:2017}: $9.4\mu$K-arcmin in polarization with a beam equal to 1.2 arcmin. In this case we set the angular scale range at $50<\ell<4000$. However, the results are quite robust against this choice. 
%since the neglected scales contribute only moderately to the reconstruction of the CMB lensing potential.

@@

\begin{figure}[htbp]
\begin{center}
\includegraphics[scale=1.]{../images/actual_scenario_planck.pdf}
\caption{Correlation factor $\rho$ with the CMB lensing potential as a function of the angular scale $\ell$ for completed and ongoing Stage-2 tracers.
Here we show both current galaxy survey and internally reconstructed CMB lensing potential. 
The dashed curve highlights which scales contribute to the $\ell<100$ B-mode power and are thus most useful to delens. It corresponds to (with arbitrary scale) $C^{\kappa \kappa} \times \frac{\partial C^{BB}}{\partial{C^{\kappa \kappa}}}$}
\label{fig:corrnow-planck}
\end{center}
\end{figure}

\begin{figure}[htbp]
\begin{center}
\includegraphics[scale=1.]{../images/actual_scenario.pdf}
\caption{Correlation factor $\rho$ with the CMB lensing potential as a function of the angular scale $\ell$ for completed and ongoing Stage-2 tracers. 
This figure is the same as \reffig{corrnow-planck} but using SPTPol instead of Planck as the high-res CMB experiment performing the internal reconstruction.
}
\label{fig:corrnow}
\end{center}
\end{figure}

Using these correlation levels we can  compute the residual B-mode power after delensing using \refeq{Bres}, and test the consequent improvement on parameter constraints with \refeq{fisher}. 
For current surveys we only report the improvements in $r$ since the broad beam of the deep BK-like survey does not provide enough leverage to constrain the shape of the B-mode spectra. We still let $n_{t}$ to vary in our fisher calculation even if it is not very degenerate with $r$.
We first compute the improvement with a fiducial value of $r_{\rm{fid}}=0$.
The results are summarized in \reftab{current} and \reftab{current-spt} for the Planck and SPTPol case respectively.

The lensing reconstruction noise in Planck is such that internal delensing can only remove $8\%$ of the power, improving the constraints on $r$ by a factor of 1.06 even in the ideal scenario of no instrumental noise in the B-modes. A similar performance is achieved with WISE. DES can do better, removing a level of 14\% of power. However, it covers a smaller fraction of the sky than the two previous probes. As shown in previous works \cite{sherwin:2015}, the best tracer is the CIB with a 27\% reduction in power.
Nonetheless, it is worthwhile combining all the tracers. Indeed a multi tracer-approach can bring the removed power on the overlapping area from 27\% of the CIB alone to 36\% using all the LSS (LSS-S2) and 41\% when Planck lensing reconstruction is added to the set.
However, if we consider a realistic level of noise in the measured B-modes, this level of residual power $C_{{l}}^{BB,\mathrm{res}}$ only leads to a 30\% improvement in $\sigma(r)$ for the null hypothesis case of $r=0$. Furthermore, as explained in \refsec{fisher}, this is only an upper limit and will be even smaller when, for example, non-perfect foreground cleaning is considered. Indeed current experiments are still significantly limited by foregrounds and instrumental noise together and not only by lensing.

@@

\begin{table}
\caption{$\alpha(r)$: Improvements on $\sigma(r)$ due to delensing for completed and ongoing Stage-2 surveys and Planck lensing reconstruction.
LSS-S2 corresponds to the combination of all the available LSS tracers.
The values in parenthesis in the first column correspond to the fraction of lensing B-mode power removed using each LSS tracer. 
The values in the other columns correspond to ratio of the error before and after delensing for 3 cases: no instrumental noise in the B-mode measurement and realistic noise for two different values of $r$.}
  \vspace{0.2cm}
  \begin{tabular}{|c | c | c | c|}
\hline
Surveys & $\alpha_{r=0}, N^{B}_{\ell}=0$ & $\alpha_{r=0}$ & $\alpha_{r=0.12}$ \\ \hline \hline
WISE (6\%)& 1.09& 1.05 & 1.01 \\ \hline
DES (14\%) &1.2 & 1.10  & 1.03 \\ \hline
CIB (27\%) &1.45& 1.21  & 1.07\\ \hline
LSS-S2 (36\%) &1.7& 1.29  & 1.09\\ \hline
CMB Planck (8\%) &1.06& 1.05  & 1.01 \\ \hline
LSS-S2 +CMB (41\%)& 1.77& 1.31& 1.10  \\ \hline
\end{tabular}
\label{tab:current}
\end{table}

@@

 \begin{table}
\caption{$\alpha(r)$: Stage-2 improvements on $\sigma(r)$ due to delensing. We use the same LSS tracers as \reftab{current} but with the internal lensing reconstruction performed by SPTPol.}
  \vspace{0.2cm}
  \begin{tabular}{|c | c | c | c|}
\hline
Surveys &$\alpha_{r=0}, N^{B}_{\ell}=0$& $\alpha_{r=0}$ & $\alpha_{r=0.12}$ \\ \hline \hline
SPTPol (35\%) &1.46& 1.27  & 1.08 \\ \hline
LSS-S2 +SPTPol (56\%) & 2.18 & 1.57 & 1.16  \\ \hline

@@

\subsection{CMB-S3 Era}\label{sec:S3}
CMB polarization measurements are rapidly improving. Indeed the next generation of ground-based telescopes has been already deployed, and data are currently being taken.
As we did for Stage-2, we model the CMB S3 as two different overlapping experiments that will be able to combine their measurement.
The two experiments combined will have a level of noise of $3\mu$K-arcmin and a 1 arcmin beam. This level of noise is also assumed for the internal noise reconstruction.
With this experimental setup, we can push the angular scale range used to calculate the Fisher matrix to $50<\ell<3000$ even if most of the high $\ell$ scale do not contribute to the constraints.
Not only will CMB experiments improve shortly: eventually DESI will start taking data. For this reason, we add DESI to the LSS tracers used in \refsec{del_res}.

@@

\begin{figure}[htbp]
\begin{center}
\includegraphics[scale=1.]{../images/S3_scenario.pdf}
\caption{Correlation factor $\rho$ with the CMB lensing potential as a function of the angular scale $\ell$. This figure is the same as \reffig{corrnow} but with a stage-3 CMB experiment performing the internal reconstrcution and with DESI added to the stage-2 galaxy surveys (LSS-S2). }
\label{fig:corrS3}
\end{center}
\end{figure}

@@

\subsection{CMB-S4 Era}\label{sec:S4}
Having as a major goal the detection of inflationary B-modes, CMB data will continue improving even after Stage-3.
An ambitious program for a Stage-4 ground CMB experiment is in the planning phase \cite{abazajian:2016}. 
Moreover, satellite and balloon CMB experiments have been proposed and have the potential to extend the accessible B-mode measurements to the largest scales. Given the unprecedented low level of noise of these experiment, delensing will be even more important than previous generations.
Following \cite{abazajian:2016}, here we assume a CMB-S4 ground experiment with a level of noise of $1\mu$K-arcmin and a 1 arcmin beam.
Furthermore, by the mid-2020s, several next-generation LSS surveys will be online. 
Here, as an example of future optical galaxy surveys, we add LSST \cite{lsst-science-collaboration:2009} to the CMB lensing tracers. 
We do not consider other surveys but we test that other experiments like Euclid \cite{laureijs:2011} and WFIRST \cite{spergel:2013} have similar delensing performance. 
Even if these LSS surveys will observe an overlapping part of the sky and similar modes, because of their different strengths and weaknesses, it will still be important to combine them efficiently.
We also consider a radio continuum survey modeled following the SKA specifications. Radio continuum observations of the 21 cm line are in their early stage, and several experimental and data-analysis challenges need to be overcome. 
However, this technique has the potential to map the LSS with relatively low noise up to redshift $z=6$. Here we considered just two possible detection thresholds at 1 GHz (flux cut): 10 $\mu$Jy and 0.1 $\mu$Jy, which are representatives of the upper and lower limit levels for a SKA phase 1 (SKA1) and a following SKA phase 2 (SKA2) survey.
Given the importance of delensing for Stage-4 experiments, here we combine radio and optical survey to test their delensing efficiency.

The correlation factor with CMB lensing for Stage-4 experiments is shown in \reffig{corrs4_10}  and in \reffig{corrs4_01} for SKA1 and SKA2 respectively.
%=======
\begin{figure}[htbp]
\begin{center}
\includegraphics[scale=1.]{../images/S4_scenario_ska10.pdf}
\caption{Correlation factor $\rho$ with the CMB lensing potential as a function of the angular scale $\ell$. This figure is the same as \reffig{corrnow} but with a stage-4 CMB experiment performing the internal reconstrcution and with LSST and SKA added to the stage-2 and stage-3 galaxy surveys (LSS-S3). }
\label{fig:corrs4_10}
\end{center}
\end{figure}
%=======
\begin{figure}[htbp]
\begin{center}
\includegraphics[scale=1.]{../images/S4_scenario_ska01.pdf}
\caption{Correlation factor $\rho$ with the CMB lensing potential as a function of the angular scale $\ell$. This figure is the same as \reffig{corrnow} but with a stage-4 CMB experiment performing the internal reconstrcution and with LSST and SKA added to the stage-2 and stage-3 galaxy surveys (LSS-S3). }
\label{fig:corrs4_01}
\end{center}
\end{figure}
%=======
A tomographically binned LSST-like experiment will be a very efficient CMB lensing tracer. Indeed it will be more than 70\% correlated with CMB lensing for all the scales $\ell<600$, a performance similar to a Stage-3 CMB internal reconstruction.
A very similar level of efficiency will be achieved by a SKA1-like experiment.
With SKA and LSST, for the first time we will have LSS tracers with a higher delensing efficiency than the CIB.
Despite the improvement in galaxy surveys, the CMB internal reconstruction will still be the main source of delensing. Having a perfect kernel overlap with the true lensing potential it will benefit from the very low level of noise of CMB S4 experiments.
\reffig{corrs4_10} shows that a CMB-S4 experiment will be able to internally reconstruct the lensing potential at more than 90\% up to $\ell=400$.
At this level, CMB S4 internal delensing will not be limited by noise but by small secondary effects like foregrounds contamination, filtered modes etc.
Finally, \reffig{corrs4_01} shows that a very futuristic SKA-2 experiment will have a very high correlation with CMB lensing, being more than 80\% correlated at all the scales. This performance is comparable to a CMB-S4 internal reconstruction.

@@

%=================================================================
\section{Acknowledgments}
We thank S. Dodelson, S. Passaglia, K. Story, A. Vieregg and K. Wu for useful discussion and their review of the draft. 
Furthermore we thank D. Alonso for his feedback on radio continuum surveys.
